---
title: "ncRNA Heart Transplantation"
output:
  html_notebook: default
---
## Setup & Libraries ##
```{r}
library(ggplot2)
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(EnhancedVolcano)
library(vsn)
library(ggvenn)
library(patchwork)
library(ggplotify)
library(ggrepel)
```
## Data Import ##
**Importing Clinical Data**
```{r}
setwd("~/GitHub/ncRNA_heart_transplant")
coldata <- read_tsv("data/phenotable.txt", show_col_types = FALSE)

#coldata <- coldata %>%
#dplyr::filter(!sample %in% c(
#    "182_S25_R1_001"
#  ))
coldata
```
**Importing Raw Counts**
```{r}
counts <- read.csv("data/miRNA_raw_counts_all_samples.csv", header = TRUE, sep = ",")
counts <- column_to_rownames(counts, var = "X")
counts <- t(counts)
rownames(counts) <- gsub("\\.", "-", rownames(counts))
head(counts)
```
**Importing Annotation Data:**
```{r}
anno <- read.csv("data/anno_report.csv", header = TRUE, sep = ",")
anno
```
**Merging & Summarization**
Calculating the sum of reads for each RNA type across every clinical condition (NR, ACR, AMR, CAV)
```{r}
merged_df <- left_join(anno, coldata, by = c("Sample.name.s." = "sample"))
summed_anno <- merged_df %>%
  filter(!is.na(condition)) %>%   
  group_by(condition) %>%
  summarise(across(where(is.numeric), sum), .groups = "drop") 

summed_anno
```
**Total read counts stratified by RNA type for each sample, Figure 1A**
Generating a stacked bar plot of the absolute read counts (in millions) across various RNA biotypes for each sample, faceted by clinical group.
```{r}
library(showtext)
font_add("Tahoma", regular = "tahoma.ttf")
showtext_auto()

merged_df$Sample.name.s. <- gsub("_.*", "", merged_df$Sample.name.s.)
anno_long <- merged_df %>%
  pivot_longer(
    cols = -c(Sample.name.s., condition, antibody), 
    names_to = "RNA_Class",
    values_to = "Count"
  )

rna_labels <- c(
  "Filtered.miRNA.Reads" = "Filtered miRNA",
  "Hairpin.miRNAs"       = "Hairpin miRNA",
  "mature.tRNA.Reads"    = "Mature tRNA",
  "mRNA.Reads"           = "mRNA",
  "ncRNA.others"         = "Other ncRNA",
  "primary.tRNA.Reads"   = "Primary tRNA",
  "rRNA.Reads"           = "rRNA",
  "snoRNA.Reads"         = "snoRNA"
)

plot_bars <- ggplot(anno_long, aes(x = as.character(Sample.name.s.), y = Count / 1000000, fill = RNA_Class)) +
  geom_col(width = 0.85) + 
  facet_grid(~ condition, scales = "free_x", space = "free_x") +
  labs(x = "Sample ID", y = "Read Count (Millions)") +
  
  scale_fill_brewer(palette = "Set3", name = "RNA Class", labels = rna_labels) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_bw(base_family = "Tahoma", base_size = 7) +
  
  theme(
    text = element_text(family = "Tahoma"),
    strip.background = element_rect(fill = "white", color = "white", linewidth = 2),
    strip.text = element_text(face = "plain", size = 8, color = "black"), 
    
    axis.line = element_line(color = "black", linewidth = 0.3),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    axis.ticks.length = unit(1, "mm"),
    axis.text = element_text(family = "Tahoma", color = "black"),
    panel.border = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing = unit(4, "mm"),
    legend.key.size = unit(4, "mm")
  )

print(plot_bars)

ggsave("./pictures_transpl/plot_distr.png", plot = plot_bars, 
       width = 168, height = 110, units = "mm", dpi = 500, bg = "white")
```


**Additional calculations**
Computing the total sum of reads for each sample
```{r}
cols <- names(anno) != "Sample.name.s."

anno["total", cols] <- colSums(anno[, cols], na.rm = TRUE)
anno["total", "Sample.name.s."] <- "total"
anno
```
**Percentage Distribution of RNA Biotypes**
Calculating the percentage of each small RNA class relative to the total number of reads in the entire experiment.
```{r}
total_vec <- anno["total", setdiff(names(anno), "Sample.name.s.")]
total_pct <- total_vec / sum(total_vec) * 100
total_pct
```
**Sequencing Depth per Group**
Aggregating the total number of reads for each clinical group (used to describe cohort characteristics in the manuscript text).
```{r}
summed_anno_plt <- summed_anno %>%
  pivot_longer(cols = -condition, names_to = "RNA_Type", values_to = "Count")

reads_before <- summed_anno_plt %>%
  group_by(condition) %>%        
  summarise(total_reads = sum(Count))
reads_before
```
## Differential Expression Analysis ##
**Custom Functions for DESeq2 Automation**
Creating two custom functions:
_run_deseq_pipeline_ — sets the reference group, applies a low-count filter (minimum of 10 reads in at least 3 samples), and runs the DESeq2 model.
_extract_and_plot_ — extracts comparison results, applies apeglm shrinkage for LFC correction, filters significant miRNAs, and generates a Volcano Plot.
```{r}
run_deseq_pipeline <- function(counts_mat, metadata, design_formula, ref_level) {
  metadata$condition <- relevel(factor(metadata$condition), ref = ref_level)
  
  dds <- DESeqDataSetFromMatrix(
    countData = counts_mat,
    colData = metadata,
    design = design_formula
  )
  
  keep <- rowSums(counts(dds) >= 10) >= 3
  dds <- dds[keep, ]
  
  dds <- DESeq(dds, fitType = "parametric")
  
  return(dds)
}

extract_and_plot <- function(dds, target_coef, plot_title, save_path) {
  res <- results(dds, name = target_coef)
  resLFC <- lfcShrink(dds, coef = target_coef, type = "apeglm")
  
  res_sign <- as.data.frame(subset(res, padj < 0.05 & !is.na(padj) & abs(log2FoldChange) > 1))
  resLFC_sign <- as.data.frame(subset(resLFC, padj < 0.05 & !is.na(padj) & abs(log2FoldChange) > 1))
  
  plt <- EnhancedVolcano(
    res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'padj',
    subtitle = NULL,  
    caption = NULL,
    legendLabels = c('NS', 'Log2 FC', 'P-adjusted', 'P-adjusted and Log2 FC'),
    ylab = bquote(~-Log[10]~ 'P-adjusted'),
    pCutoff = 0.05,
    FCcutoff = 1.0,
    labSize = 3.0, 
    boxedLabels = FALSE,
    col = c('black', '#CBD5E8', '#B3E2CD', "#FF6600"),
    colAlpha = 1,
    title = NULL,
    selectLab = rownames(res_sign),  
    drawConnectors = TRUE,
    xlim = c(-10, 10),  
    ylim = c(0, 5)
)
  
  plt <- plt + 
    theme_bw(base_family = "Tahoma", base_size = 7)
  ggsave(save_path, plot = plt, width = 8, height = 6, dpi = 500, bg = "white")
  
  return(list(
    res_raw = res,
    res_shrunken = resLFC,
    sig_raw = res_sign,
    sig_shrunken = resLFC_sign,
    plot = plt
  ))
}
```
### Pairwise Comparisons with the Normal Reference (NR as Reference) ###
**Differential expression analysis for Antibody-Mediated Rejection (AMR vs NR), Figure 2A**
```{r}
dds_nr_ref <- run_deseq_pipeline(counts, coldata, ~ condition, "NR")

# AMR vs NR
res_AMR_vs_NR <- extract_and_plot(
  dds = dds_nr_ref, 
  target_coef = "condition_AMR_vs_NR", 
  plot_title = "AMR vs NR", 
  save_path = "./pictures_transpl/transpl_VolcanoPlot_AMR_vs_NR.svg"
)
res_AMR_vs_NR$plot
```
```{r}
res_AMR_vs_NR$sig_raw
```
```{r}
res_AMR_vs_NR$sig_shrunken 
```
**Per-sample normalized expression of the principal candidate miRNAs across clinical groups, Figure 3**
```{r}
target_mirnas <- c("hsa-miR-193b-5p", "hsa-miR-125b-5p", "hsa-miR-5588-5p")
all_counts <- data.frame()

for (mirna in target_mirnas) {
  count_data <- plotCounts(dds_nr_ref, gene=mirna, intgroup="condition", returnData=TRUE)
  count_data$sample <- gsub("_.*", "", rownames(count_data))
  count_data$miRNA <- mirna 
  all_counts <- rbind(all_counts, count_data)
}

safe_colors <- c(ACR = "#E69F00", AMR = "#56B4E9", CAV = "#CC79A7", NR = "#999999") 

plt <- ggplot(all_counts, aes(x=condition, y=count, fill=condition)) +
  geom_boxplot(alpha=0.6, outlier.shape = NA, color="black", linewidth=0.3) + 
  
  geom_jitter(width=0.2, size=1.2, alpha=0.8, shape=21, color="black") +
  
  geom_text_repel(aes(label=sample), size=2, max.overlaps=15) + 
  
  scale_y_log10() +
  scale_fill_manual(values = safe_colors) + 
  facet_wrap(~ miRNA, scales="free_y", ncol=3) + 
  
  labs(y = "Normalized expression counts", x = "Clinical group") + 
  theme_classic(base_family = "Tahoma", base_size = 7) + 
  
  theme(
    strip.text = element_text(size = 7, face = "bold.italic"), 
    strip.background = element_rect(fill = NA, color = NA),
    legend.position = "none",
    
    axis.line = element_line(color = "black", linewidth = 0.3),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    axis.ticks.length = unit(1, "mm"),
    axis.text = element_text(color = "black"),
    panel.spacing = unit(4, "mm") 
  )

print(plt)
# Statistical graphs should have 300-500 dpi[cite: 1].
# Figures should be submitted in lossless PNG format[cite: 1].
ggsave("./pictures_transpl/126281-g003.png", plot = plt, width = 168, height = 75, units = "mm", dpi = 500)
ggsave("./pictures_transpl/126281-g003.svg", plot = plt, width = 168, height = 75, units = "mm")
```


**Differential expression analysis for Acute Cellular Rejection (ACR vs NR)**
```{r}
res_ACR_vs_NR <- extract_and_plot(
  dds = dds_nr_ref, 
  target_coef = "condition_ACR_vs_NR", 
  plot_title = "ACR vs NR", 
  save_path = "./pictures_transpl/transpl_VolcanoPlot_ACR_vs_NR.svg"
)
res_ACR_vs_NR$sig_raw
res_ACR_vs_NR$sig_shrunken 
```
**Differential expression analysis for Cardiac Allograft Vasculopathy (CAV vs NR).**
```{r}
res_CAV_vs_NR <- extract_and_plot(
  dds = dds_nr_ref, 
  target_coef = "condition_CAV_vs_NR", 
  plot_title = "CAV vs NR", 
  save_path = "./pictures_transpl/transpl_VolcanoPlot_CAV_vs_NR.svg"
)
res_CAV_vs_NR$plot
```
```{r}
res_CAV_vs_NR$sig_raw
```
```{r}
res_CAV_vs_NR$sig_shrunken 
```
### Comparisons Between Rejection Types (ACR as Reference) ###
**Changing the baseline to ACR to identify differences between rejection types (AMR vs ACR), Figure 2D**
```{r}
dds_acr_ref <- run_deseq_pipeline(counts, coldata, ~ condition, "ACR")

res_AMR_vs_ACR <- extract_and_plot(
  dds = dds_acr_ref, 
  target_coef = "condition_AMR_vs_ACR", 
  plot_title = "AMR vs ACR", 
  save_path = "./pictures_transpl/transpl_VolcanoPlot_AMR_vs_ACR.svg"
)
res_AMR_vs_ACR$plot
```
```{r}
res_AMR_vs_ACR$sig_raw
```
```{r}
res_AMR_vs_ACR$sig_shrunken
```
**Changing the baseline to ACR to identify differences between rejection types (CAV vs ACR)**
```{r}
res_CAV_vs_ACR <- extract_and_plot(
  dds = dds_acr_ref, 
  target_coef = "condition_CAV_vs_ACR", 
  plot_title = "CAV vs ACR", 
  save_path = "./pictures_transpl/transpl_VolcanoPlot_CAV_vs_ACR.svg"
)
res_CAV_vs_ACR$sig_raw
res_CAV_vs_ACR$sig_shrunken 
```
### Pairwise Comparisons with CAV (CAV as Reference) ###
**Differential expression analysis for Antibody-Mediated Rejection (AMR vs NR), Figure 2B**
```{r}
dds_cav_ref <- run_deseq_pipeline(counts, coldata, ~ condition, "CAV")

res_AMR_vs_CAV <- extract_and_plot(
  dds = dds_cav_ref, 
  target_coef = "condition_AMR_vs_CAV", 
  plot_title = "AMR vs CAV", 
  save_path = "./pictures_transpl/transpl_VolcanoPlot_AMR_vs_CAV.svg"
)
res_AMR_vs_CAV$plot
```
```{r}
res_AMR_vs_CAV$sig_raw
```
```{r}
res_AMR_vs_CAV$sig_shrunken 
```
**Changing the baseline to CAV to evaluate ACR vs CAV**
```{r}
res_ACR_vs_CAV <- extract_and_plot(
  dds = dds_cav_ref, 
  target_coef = "condition_ACR_vs_CAV", 
  plot_title = "ACR vs CAV", 
  save_path = "./pictures_transpl/transpl_VolcanoPlot_ACR_vs_CAV.svg"
)
res_ACR_vs_CAV$sig_raw
res_ACR_vs_CAV$sig_shrunken 
```
### Covariate-Adjusted Model ###
**Adding antibody status (antibody) to the model design (~ antibody + condition) to account for the potential confounding effect of HLA sensitization on the expression profile, Figure 2C**
```{r}
dds_covariates <- run_deseq_pipeline(counts, coldata, ~ antibody + condition, "NR")

res_AMR_vs_NR_adj <- extract_and_plot(
  dds = dds_covariates, 
  target_coef = "condition_AMR_vs_NR", 
  plot_title = "AMR vs NR (Adjusted for HLA)", 
  save_path = "./pictures_transpl/transpl_VolcanoPlot_AMR_vs_NR_adjusted.svg"
)
res_AMR_vs_NR_adj$plot
```
```{r}
res_AMR_vs_NR_adj$sig_raw
```
```{r}
res_AMR_vs_NR_adj$sig_shrunken
```
## Venn Diagram ##
**Venn diagram showing the overlap of filtered miRNAs among the clinical groups, Figure 1D**
Applying a global noise filter to the raw counts and identifying the lists of miRNAs confidently expressed in each clinical group to build an intersection diagram
```{r}
keep_global <- rowSums(counts >= 10) >= 3
filtered_counts <- counts[keep_global, ]

counts_mat <- filtered_counts

get_filtered_mirnas <- function(condition_name){
    samps <- coldata %>%
        filter(condition == condition_name) %>%
        pull(sample)
    valid_samps <- intersect(samps, colnames(counts_mat))
    sub_counts <- counts_mat[, valid_samps, drop = FALSE]
    keep_group <- rowSums(sub_counts > 0) > 0
    rownames(sub_counts)[keep_group]
}

venn_list <- list(
  NR  = get_filtered_mirnas("NR"),
  ACR = get_filtered_mirnas("ACR"),
  AMR = get_filtered_mirnas("AMR"),
  CAV = get_filtered_mirnas("CAV")
)

plot_venn <- ggvenn(
  venn_list,
  fill_color = c("#999999", "#E69F00", "#56B4E9", "#CC79A7"),
  fill_alpha = 0.5,
  stroke_size = 0,
  
  set_name_size = 3, 
  text_size = 2      
)

plot_venn <- plot_venn + 
  theme_void(base_family = "Tahoma")

print(plot_venn)
ggsave("./pictures_transpl/venna_plot.svg", plot = plot_venn, width = 8, height = 6, dpi = 500, bg = "white")
```
## Unsupervised Clustering & PCA ##
**Applying a regularized logarithm (rlog) transformation to the DESeq2 object to stabilize the variance prior to clustering methods. The meanSdPlot confirms the removal of variance dependence on the mean**
```{r}
rlt <- rlog(dds_nr_ref)
meanSdPlot(assay(rlt)) 
```
## Principal Component Analysis (PCA) ##
**Principal component analysis (PCA) of filtered miRNA expression profiles according to clinical group and the presence of contemporaneous anti-HLAs, Figure 1B**
Generating a PCA plot to evaluate the global segregation of samples in a 2D space based on their expression profiles. Colors encode the clinical group, and shapes represent the antibody status
```{r}
pcaData <- plotPCA(rlt, intgroup=c("condition", "antibody"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

pcaData$sample <- gsub("_.*", "", rownames(pcaData))
 
custom_palette <- c(ACR = "#E69F00", AMR = "#56B4E9", CAV = "#CC79A7", NR = "#999999")

plot_pca <- ggplot(pcaData, aes(PC1, PC2, color = condition, shape = antibody)) +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = sample),
    size = 2, 
    box.padding = 0.4,
    point.padding = 0.3,
    segment.color = "grey50",
    segment.size = 0.4,
    max.overlaps = Inf
  ) +
  xlab(paste0("PC1: ", percentVar[1], "%")) +
  ylab(paste0("PC2: ", percentVar[2], "%")) +
  coord_fixed() +
  
  theme_bw(base_family = "Tahoma", base_size = 7) +
  scale_color_manual(values = custom_palette) +
  

  theme(
    axis.line = element_line(color = "black", linewidth = 0.3),
    axis.ticks = element_line(color = "black", linewidth = 0.3),
    axis.ticks.length = unit(1, "mm"),
    axis.text = element_text(color = "black"),
    
    legend.key.size = unit(3, "mm"),
    legend.title = element_text(size = 7, face = "bold"),
    legend.text = element_text(size = 6),
    
    panel.border = element_blank(),
    panel.grid.minor = element_blank()
  )

plot_pca
ggsave("./pictures_transpl/transpl_PCA_plot.svg", plot = plot_pca, width = 8, height = 6, dpi = 500, bg = "white")
```
## Sample Distance Heatmap ##
**Distance-based heatmap of inter-sample correlations, annotated by clinical group and contemporaneous anti-HLA antibody status, Figure 1C**
Calculating the Euclidean distance between all samples and generating a heatmap with hierarchical clustering to visualize patient similarities
```{r}
#Calculate distances and convert to matrix
sampleDists <- dist(t(assay(rlt)))
sampleDistMatrix <- as.matrix(sampleDists)

rownames(sampleDistMatrix) <- gsub("_.*", "", colnames(rlt))
colnames(sampleDistMatrix) <- gsub("_.*", "", colnames(rlt))

# Extracting the columns you mentioned using in plotPCA
anno_df <- as.data.frame(colData(rlt)[, c("condition", "antibody")])

# Capitalize the column names so the legends appear as "Condition" and "Antibody"
colnames(anno_df) <- c("Condition", "Antibody")

# Define custom colors for your annotations to match the image
anno_colors <- list(
  Condition =  c(ACR = "#E69F00", AMR = "#56B4E9", CAV = "#CC79A7", NR = "#999999"),
  Antibody = c(yes = "#FF6600", no = "#E2E2E2")
)
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255) 

plot_hm <- pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = "euclidean", 
  clustering_distance_cols = "euclidean",
  color = colors,                        
  annotation_col = anno_df,
  annotation_row = anno_df,
  annotation_colors = anno_colors,
  annotation_names_row = FALSE,
  annotation_names_col = FALSE,
  fontsize = 5,         
  fontsize_row = 4,      
  fontsize_col = 4,     
  angle_col = "45",     
  treeheight_row = 15,  
  treeheight_col = 15,   
  border_color = NA,
  cellwidth = 4.8,   
  cellheight = 4.8
)

ggsave("./pictures_transpl/transpl_distance_heatmap.svg", plot = plot_hm, width = 9, height = 6, dpi = 500, bg = "white")
```
## Likelihood Ratio Test ##
**Multi-Group Analysis (LRT). Performing a Likelihood Ratio Test comparing the full model (~ condition) against an intercept-only reduced model (~ 1) to identify miRNAs whose expression significantly changes in at least one group compared to all others.**
```{r}
dds_lrt <- DESeq(dds_nr_ref, test = "LRT", reduced = ~ 1)

res_all <- results(dds_lrt, alpha = 0.05)
summary(res_all)
```
```{r}
res_all_sign <- as.data.frame(subset(res_all, padj < 0.05 & !is.na(padj)))
res_all_sign
```
**Combined Figure 1**
```{r}
plot_bars <- plot_bars + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 5), 
    legend.key.size = unit(3, "mm"), 
    legend.text = element_text(size = 5), 
    legend.title = element_text(size = 6, face = "bold")
  )

plot_pca <- plot_pca + 
  theme(
    legend.key.size = unit(3, "mm"),
    legend.text = element_text(size = 5)
  )

plot_hm <- as.ggplot(plot_hm) 

combined_figure <- (plot_bars | plot_pca) / 
                   (plot_hm | plot_venn)

final_figure <- combined_figure + 
  plot_annotation(tag_levels = 'A') & 
  theme(
    plot.tag = element_text(family = "Tahoma", face = "bold", size = 10),
    plot.margin = margin(3, 3, 3, 3, "mm")
  )

ggsave(
  "./pictures_transpl/126281-g001.svg", 
  plot = final_figure, 
  width = 168,       
  height = 120, 
  units = "mm", 
  bg = "white"
)

ggsave(
  "./pictures_transpl/126281-g001.png", 
  plot = final_figure, 
  width = 168,       
  height = 120,
  units = "mm", 
  dpi = 600, 
  bg = "white"
)
final_figure
```
**Combined Figure 2**
```{r}
combined_figure <- (res_AMR_vs_NR$plot | res_AMR_vs_CAV$plot) / 
                   (res_AMR_vs_NR_adj$plot | res_AMR_vs_ACR$plot) + 

  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A') & 
  
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    plot.tag = element_text(family = "Tahoma", face = "bold", size = 10),
    plot.margin = margin(5, 5, 5, 5, "mm")
  )

ggsave(
  "./pictures_transpl/126281-g002.png", 
  plot = combined_figure, 
  width = 168, 
  height = 200, 
  units = "mm", 
  dpi = 600, 
  bg = "white"
)
ggsave(
  "./pictures_transpl/126281-g002.svg", 
  plot = combined_figure, 
  width = 168, 
  height = 200, 
  units = "mm", 
  dpi = 600, 
  bg = "white"
)
combined_figure
```


